Continuous gravitational waves in the lab: recovering audio signals with a table-top optical microphone

### 2020-09-09 paper/ifo-introduction.tex

% general gw intro
In 2015, gravitational waves were observed for the first time from the merger of two black holes~\cite{GW150914}. 
The observation, made by the Laser Interferometer Gravitational-wave Observatory~\citep[LIGO]{AdvancedLIGO:2015}, marked a breakthrough in modern astrophysics and revealed a new means to observe the universe. 
Since 2015, the LIGO and Virgo~\cite{AdvancedVirgo:2015} observatories have made numerous detections of binary black hole~\cite{GW151226,GW170104,GW170814} and binary neutron star~\cite{GW170817,GW170817multi,GW190425} mergers. 
Gravitational-wave detection has led to increased public and media interest in gravitational-wave science. 
Gravitational-wave research groups around the world have produced demonstrations and activities to explain their research to general audiences. 
Activities range from hands-on demonstrations, exhibitions~\cite{TTExhibit:2020,LIGOScienceEducationCentre:online,GravityDiscoveryCentre:online}, and online data analysis tutorials~\cite{GWOSC:online,LOSC:2015}, to phone apps~\cite{LaserLabs:online,SciVR:online,chirp:online}, online games~\cite{BlackHoleHunter:online}, and musical performances~\cite{ArthurJeffesMusic:online,GravitySynthLeonTrimble:online}%\han{look up Sarah Farmer's project}.

% what are gws and how do we detect them...
Gravitational waves are a prediction of Albert Einstein's theory of general relativity~\cite{Einstein:1916}. 
They are disturbances in spacetime caused by the acceleration of asymmetric massive objects. 
The effect of gravitational waves is a change in lengths: a `stretching and squeezing' of the distance between objects.
Observatories such as LIGO, Virgo, and KAGRA~\cite{KAGRA:2013} are laser interferometers.
They use the interference of laser light to measure changes in distance. 
%As shown in an aerial view of LIGO Hanford in Figure~\ref{fig:ligo_pic}, they use beam arms kilometres long.
Gravitational-wave observatories are extremely complex but are fundamentally based on the Michelson interferometer. 
Table-top Michelson interferometers are commonly used in undergraduate lab experiments~\cite{UgoliniEtAl:2019} and to demonstrate the science of gravitational wave detection to non-specialist audiences~\cite{ThorLabsIFO,NikhefIFO,TTExhibit:2020,LIGOIFOGlue,LIGOIFOMagnets}.

% continuous gws
To date the network of gravitational wave observatories has observed short-duration transient signals~\cite{GWTC-1:2018,GWOSC:online}. 
However, the network is also searching for continuous gravitational waves: persistent, periodic, near-monochromatic signals.
Continuous waves may be emitted at constant frequencies, or may wander slowly in frequency over time. 
For example, rotating neutron stars in low mass X-ray binaries (LMXBs) are one of the targets for continuous-wave searches. 
LMXBs are binary systems containing a compact object (such as a neutron star or a black hole) in orbit with a low mass stellar companion, where the rotation frequency of the compact object wanders over time due to variable accretion of matter (and hence angular momentum transfer) from the companion~\cite{xraybinaries:1997}. 
Scorpius X-1 is an LMXB which is particularly bright in X-rays. 
It is a prime target for continuous wave searches from LMXBs and numerous, as yet unsuccessful, searches have been performed~\cite{ScoX1O2Viterbi:2019,RadiometerO1O2:2019,SearchCrossCorrO1:2017}.

% sound analogy
In this paper, we present a table-top experiment, where a standard Michelson interferometer plays the role of a `toy' gravitational-wave detector, which detects acoustic vibrations (such as music) instead of gravitational waves. 
As an undergraduate lab experiment this demonstration can be used to teach topics ranging from gravitational-wave detection and analysis to electronics, signal processing, and speech enhancement. 
It gives students in courses such as physics and electrical engineering the freedom to explore the response of an accessible yet nontrivial opto-mechanical system using a hierarchy of data analysis techniques of increasing complexity, including those used in the search for continuous waves in LIGO and Virgo data~\cite{ScoX1O2Viterbi:2019,ScoX1ViterbiO1:2017,SuvorovaEtAl:2017,SuvorovaEtAl:2017}.
The demonstration can also be adapted for use as an outreach tool to explain gravitational-wave science to a general audience.

The paper presents 
In this work we present a selection of demonstrations using different types of audio signals ranging from simple tones, analogous to continuous wave signals, to more complex audio of music and speech. 
In Section~\ref{sec:ifo} we detail the table-top interferometer design. 
We demonstrate observing a single note from a speaker in Section~\ref{sec:single_tone}.
In Section~\ref{sec:viterbi_wandering} we replace the single note with a slowly changing frequency and analyse the signal using a hidden Markov model technique from continuous-wave searches. 
In Section~\ref{sec:optical_microphone}, we demonstrate capturing and playing back complex audio, such as music and speech. 
This demonstration of an `optical microphone' serves as a more general exhibition of signal processing. 
We suggest avenues of future work in Section~\ref{sec:future_work} and draw conclusions in Section~\ref{sec:conclusions}.
Appendix~\ref{app:code} collects the software and scripts used to produce this work.

>>>>

% general gw intro
In 2015, the first detection of gravitational waves from the merger of two black holes~\cite{GW150914} by the Advanced Laser Interferometer Gravitational-wave Observatory~(aLIGO \citep{AdvancedLIGO:2015}) marked a breakthrough in modern astrophysics and opened up a new window onto the universe. 
% what are gws and how do we detect them...
Gravitational waves, predicted by Albert Einstein in his theory of general relativity~\cite{Einstein:1916}, are disturbances in `the fabric of' spacetime caused by the acceleration of massive objects.
% technically incorrect, but do we need to mention antiaxial asymmetry here? 
Observatories such as aLIGO, aVirgo, GEO600~\cite{GEO600:2010}, and KAGRA~\cite{KAGRA:2013} detect these waves as tiny changes in length: a `stretching and squeezing' of the distance between objects.
In particular, they use the interference of light to measure changes in the lengths of two perpendicular arms, kilometres long.

% continuous gws
To date, the network of gravitational-wave observatories has detected numerous signals from binary black hole~\cite{GW151226,GW170104,GW170814} and binary neutron star~\cite{GW170817,GW170817multi,GW190425} mergers, short-duration transient signals~\cite{GWTC-1:2018,GWOSC:online}. 
However, the network is also searching for continuous gravitational waves: persistent, periodic, near-monochromatic signals, but is as yet unsuccessful~\cite{ScoX1O2Viterbi:2019,RadiometerO1O2:2019,SearchCrossCorrO1:2017,ScoX1ViterbiO1:2017,SearchRadiometerO1:2017,SearchCrossCorrO1:2017,MiddletonEtAlO2LMXBs:2020,SearchTwoSpecS6:2017}.
Predicted sources for such continuous (gravitational) waves include rotating neutron stars in orbit around low mass stellar companions (e.g.\ low mass X-ray binaries, LMXBs). Continuous waves from such sources may be emitted at constant frequency or may wander slowly in frequency over time. 
For example, the rotation frequency of a neutron star in a LMXB can wander over time due to variable accretion of matter (and hence angular momentum transfer) from the stellar companion~\cite{xraybinaries:1997}, and this rotation frequency is predicted to determine the frequency of the continuous waves.
The prime target for continuous wave searches from LMXBs is Scorpius X-1, an LMXB particularly bright in X-rays.

% table-top
Gravitational-wave observatories are extremely complex but are fundamentally based on the classical Michelson interferometer. 
Table-top Michelson interferometers are commonly used in undergraduate laboratory experiments~\cite{UgoliniEtAl:2019} and to demonstrate the science of gravitational wave detection to non-specialist audiences~\cite{ThorLabsIFO,NikhefIFO,AMIGO:online,TTExhibit:2020,LIGOIFOGlue,LIGOIFOMagnets}.
% sound analogy
In this paper, we use a table-top Michelson interferometer as a `toy' gravitational-wave observatory designed to detect sound instead of gravitational waves. That is, instead of detecting changes in the spacetime along the arms, we detect Newtonian motion of the mirrors driven by acoustic vibrations in the air around them. 
The design of the interferometer is detailed in Section~\ref{sec:ifo}.
By playing different sounds through a speaker aimed at the interferometer, capturing the resultant interference pattern, and applying suitable signal processing, we can play-back the original signal and as such turn the interferometer into an `optical microphone', using light to capture sound.
We observe a constant frequency signal in Section~\ref{sec:single_tone}, a wandering frequency signal with application of the Viterbi algorithm (a hidden Markov model technique) from continuous wave searches in Section~\ref{sec:viterbi_wandering}, and more complex audio, such as music and speech, in Section~\ref{sec:optical_microphone}, to varying degrees of success. The latter being less about applying gravitation wave techniques and more about exhibiting general signal processing.
We suggest avenues of future work in Section~\ref{sec:future_work} and draw conclusions in Section~\ref{sec:conclusions}.
% Appendix~\ref{app:code} collects the software and scripts used to produce this work.

% broader future applications
This demonstration can be used to teach topics ranging from gravitational-wave detection and analysis to electronics, signal processing, and speech enhancement. 
Additionally, it allows students in physics and electrical engineering courses to explore the response of an accessible, yet nontrivial, opto-mechanical system using a hierarchy of data analysis techniques of increasing complexity, up to those used in the search for continuous waves in aLIGO and aVirgo data~\cite{ScoX1O2Viterbi:2019,ScoX1ViterbiO1:2017,SuvorovaEtAl:2017,SuvorovaEtAl:2017}.

Moreover, gravitational-wave detection has led to increased public and media interest in gravitational-wave science. 
Research groups around the world have produced demonstrations and activities to explain their research to general audiences, with efforts ranging from hands-on demonstrations, exhibitions~\cite{TTExhibit:2020,LIGOScienceEducationCentre:online,GravityDiscoveryCentre:online}, and online data analysis tutorials~\cite{GWOSC:online,LOSC:2015}, to phone applications~\cite{LaserLabs:online,SciVR:online,chirp:online}, online games~\cite{BlackHoleHunter:online}, and musical performances~\cite{ArthurJeffesMusic:online,GravitySynthLeonTrimble:online}.%\han{look up Sarah Farmer's project}
This demonstration can be adapted to an effective outreach tool to join those above in explaining gravitational wave science to a non-specialist audience.

### 2020-09-09 paper/paper_main.tex

\begin{abstract}
Gravitational-wave observatories around the world are searching for continuous waves: persistent signals from spinning neutron stars. 
Searches use sophisticated statistical techniques to look for weak signals in noisy data. 
We show how continuous-wave searches can be demonstrated with a table-top model gravitational-wave detector (a Michelson interferometer), where sound plays the role of gravitational waves. 
Using signal processing techniques from continuous-wave searches we demonstrate the recovery of tones with constant and wandering frequencies. 
Finally we investigate the performance of the demonstration as an `optical microphone' for recovering music and speech. 
Simple chords and drums are easily recovered, but speech is more challenging. 
This demonstration has scope for use in the physics and electrical engineering undergraduate lab. 
It can also be adapted for use as an engagement tool for communicating gravitational-wave and signal-processing topics to non-specialist audiences. 
\end{abstract}

>>>>

\begin{abstract}
Gravitational-wave observatories around the world are searching for continuous waves: persistent signals from spinning neutron stars. 
These searches use sophisticated statistical techniques to look for weak signals in noisy data. 
In this paper, we demonstrate these techniques using a table-top model gravitational-wave detector (a Michelson interferometer), where sound plays the role of gravitational waves. 
Using signal processing techniques from continuous-wave searches we demonstrate the recovery of tones with constant and wandering frequencies. 
Finally we use the interferometer as an `optical microphone' to capture music and speech and apply filtering techniques to recover the original signals from the noisy data. Such techniques include simple bandpass and notch filters, as well as comparing a Weiner filter to the statistical logMMSE method.  
Simple chords and drums are easily recovered, but complex music and speech are more challenging. 
This demonstration has scope for use in physics and electrical engineering undergraduate laboratories. 
It can also be adapted for use as an engagement tool for communicating gravitational-wave and signal-processing topics to non-specialist audiences. 
\end{abstract}
